% Section 1: Introduction
\label{sec:introduction}

Artificial intelligence is rapidly becoming the infrastructure of financial markets. Machine learning methods determine which securities are cheap and which are expensive. Large language models produce the financial analyses that inform investment committees. Algorithmic systems calibrate the models that market makers use to price liquidity. This transformation is well documented: diverse machine learning methods converge on the same dominant predictors of equity returns, and large language models at fixed temperature generate near-identical financial statement analyses. The convergence is not an aberration. It is a structural feature of training on common data and deploying pre-trained models at scale.

The central question is: what happens to financial market stability when the private signals of many agents become highly correlated because they are all processed by the same AI system? The question is not about whether AI makes individual agents better informed. It is about what happens when it makes them all informed in the same way.

We formalise AI signal homogeneity through a single primitive: the cross-sectional correlation $\rho \in [0,1]$ of AI-generated signals. Each agent's signal error decomposes into a common AI model error shared by all AI-equipped agents and an idiosyncratic component, with $\rho$ governing the fraction of variance attributable to the common component. At $\rho = 0$, signals are independent and the economy sits at the Morris-Shin heterogeneous-signal benchmark. At $\rho = 1$, every AI-equipped agent observes the same signal. The entire paper is organised as comparative statics in $\rho$.

We show that rising $\rho$ activates fragility through three channels, each connected to an established literature but none sufficient on its own.

Channel 1 (coordination failure) operates through the \citet{goldstein2005} bank-run framework. As $\rho$ rises, AI signals increasingly resemble a common public signal, progressively restoring the multiplicity region of \citet{hellwig2002}. The crisis threshold $\theta^*(\rho)$ is non-monotone: declining at low $\rho$ (a ``wisdom of AI'' effect, where correlated signals improve individual inference) and rising at high $\rho$ (a ``herding of AI'' effect, where the common shock coordinates withdrawal decisions), with uniqueness failing at $\rho_1^* = 1/(1+\alpha_{\text{SC}}^2)$. The economic intuition is that moderate correlation resembles better private information, while high correlation resembles a public signal that enables self-fulfilling runs. But coordination failure alone does not explain why prices stop conveying useful information.

Channel 2 (information acquisition collapse) operates through the \citet{grossman1980} framework. The information collapse result shows that the aggregate information content of many AI-equipped agents is bounded by that of a small number of independent agents (inversely proportional to $\rho$), and revelatory price efficiency is monotonically decreasing in $\rho$ for $c_P > 0$. Forecasting price efficiency, by contrast, is non-monotone: it improves at low $\rho$ as cheap AI signals contribute diverse information, then declines once information collapse dominates. This FPE-RPE divergence means that standard forecast-accuracy metrics can paint a misleadingly favourable picture. Yet information loss alone does not explain why liquidity vanishes.

Channel 3 (correlated liquidity withdrawal) operates through the \citet{avellaneda2008} market-making framework. Shared AI calibration reduces the effective number of independent liquidity providers, $N_{\text{eff}}$, which collapses to one as $\rho \to 1$. A market with a hundred nominally independent market makers may behave as if it has fewer than ten. A no-equilibrium threshold $\rho^{**}$ exists above which market makers withdraw simultaneously. Taken individually, each channel has its own fragility threshold. Taken together, they have a lower one.

The paper's central result is that the three channels interact and amplify. We define a composite operator mapping the state vector of effective correlation, crisis threshold, and effective market-maker count into itself through three equilibrium mappings that formalise the cross-channel feedback. By Brouwer's fixed-point theorem, this operator admits at least one fixed point on a compact convex domain.

This fixed point is where the danger lies. Let $\rho_1^*, \rho_2^*, \rho_3^*$ denote the individual channel fragility thresholds and $\rho^*$ the bifurcation threshold of the integrated system. We show that
\begin{equation}\label{eq:main-result-intro}
  \rho^* < \min(\rho_1^*, \rho_2^*, \rho_3^*).
\end{equation}
The integrated system becomes fragile at a strictly lower AI signal correlation than any individual channel assessment predicts. There exists a non-empty interval $(\rho^*, \min_i \rho_i^*)$ in which every single-channel stress test returns a favourable verdict while the system has already crossed its compound fragility threshold. We call this the \emph{safety illusion}. It is the paper's central message: any regulator examining one channel at a time will miss the compound threshold entirely.

We derive five contributions, each stated as a formal proposition.

First, we characterise the crisis threshold $\theta^*(\rho)$ in the Goldstein-Pauzner bank-run game (Propositions~\ref{prop:uniqueness-boundary} and~\ref{prop:crisis-nonmonotone}). The threshold is non-monotone in $\rho$ with a uniqueness-to-multiplicity phase transition at $\rho_1^* = 1/(1+\alpha_{\text{SC}}^2)$. Below this boundary, a unique threshold equilibrium exists and comparative statics are well-defined; above it, self-fulfilling crises become possible regardless of fundamentals. The social cost of AI signal correlation is characterised in Proposition~\ref{prop:welfare-channel1} via the \citet{angeletos2007} framework: the welfare loss is linearly increasing in $\rho$, creating a wedge between private incentives and social optimum that motivates the extension in Section~\ref{sec:extensions}.

Second, we derive the cross-sectional Grossman-Stiglitz extension with two information types (Propositions~\ref{prop:info-diversity} and~\ref{prop:rpe}). Revelatory price efficiency, which measures the novel information that prices reveal beyond the common AI signal, is monotonically decreasing in $\rho$ for $c_P > 0$. Forecasting price efficiency is non-monotone: it initially improves as cheap AI signals enter the market, then declines once information collapse dominates. The divergence between the two measures means that AI can appear to improve price accuracy by standard metrics while degrading the information that matters for real investment decisions. The Goldstein-Yang complementarity mechanism breaks down at high $\rho$ (Proposition~\ref{prop:gy-breakdown}), producing an information monoculture.

Third, we derive the liquidity fragility index $N_{\text{eff}}(\rho)$, an effective-number formula that captures how correlated AI calibrations reduce independent liquidity supply. We prove its strict monotonicity and convexity in $\rho$ (Proposition~\ref{prop:neff}) and derive the no-equilibrium threshold (Proposition~\ref{prop:rho-star-star}). The convexity is economically important: the marginal loss of effective independence is greatest at low $\rho$, where diversification benefits are most valuable. The \citet{danielsson2012} common VaR framework is nested as the limiting case $\rho = 1$ (Proposition~\ref{prop:dsz}).

Fourth, and most importantly, we establish the amplification loop. We prove fixed-point existence (Proposition~\ref{prop:fixed-point-existence}), the bifurcation result~\eqref{eq:main-result-intro} (Proposition~\ref{prop:bifurcation}), local uniqueness below the threshold (Proposition~\ref{prop:local-uniqueness}), and the safety illusion (Corollary~\ref{cor:safety-illusion}). The economic content of the loop is concrete: correlated liquidity withdrawal raises the effective signal correlation faced by depositors, which raises the crisis threshold, which destroys the option value of private research, which increases market-maker reliance on the common AI signal, which further reduces $N_{\text{eff}}$. The safety illusion is the result the entire analysis builds toward.

Fifth, we endogenise $\rho$ through a prisoner's dilemma in AI adoption (Proposition~\ref{prop:endogenous-rho}): the competitive equilibrium correlation $\rho^{\text{NE}}$ exceeds the social optimum $\rho^{\text{SO}}$, and when the amplification loop is operative, $\rho^{\text{NE}} > \rho^*$. Each fund manager benefits individually from adopting the dominant AI model, but aggregate adoption raises systemic fragility. A diversity mandate bounding $\rho$ below $\rho^*$ restores efficiency (Proposition~\ref{prop:diversity-mandate}).

The closest predecessors are \citet{danielsson2025} and \citet{yang2024}. \citet{danielsson2025} derive a crisis threshold as a function of $\mu$, the fraction of AI agents, through an execution-speed channel. Their primitive captures the extensive margin of AI penetration; ours captures the intensive margin of AI output homogeneity. Moreover, \citet{danielsson2025} formally model only the coordination/run channel and do not integrate the \citet{hellwig2002} multiplicity result into their formal model; they do not model information acquisition collapse, market-maker liquidity withdrawal, or the cross-channel amplification loop. \citet{yang2024} demonstrates AI coordination computationally via Q-learning agents but has no signal correlation parameter and no analytical characterisation. \citet{hansen2025} find that independently queried AI agents reduce herding, consistent with the low-$\rho$ regime; they do not test the high-$\rho$ regime where our fragility mechanism activates.

In the information acquisition literature, \citet{dugast2018} and \citet{farboodi2020} model temporal tradeoffs in data technology but do not address the cross-sectional dimension. In market microstructure, \citet{danielsson2012} formalise procyclical selling from common VaR constraints, which we nest as the $\rho = 1$ limiting case. \citet{danielsson2022} provide the most comprehensive qualitative argument for multi-channel AI systemic risk; \citet{dou2025a, dou2025b} model AI fragility through reinforcement learning. Our mechanism is complementary, arising from signal homogeneity rather than strategic learning. Section~\ref{sec:literature} provides a full discussion.

Section~\ref{sec:model} introduces the signal structure, agent types, timing, and equilibrium concepts. Sections~\ref{sec:channel1}--\ref{sec:channel3} analyse Channels 1, 2, and 3 respectively. Each section demonstrates that its channel creates fragility in isolation; each also reveals why that channel alone provides an incomplete picture. Section~\ref{sec:amplification} formalises the amplification loop and derives the bifurcation result. Section~\ref{sec:extensions} endogenises $\rho$ and characterises the diversity mandate. Section~\ref{sec:empirics} provides motivating evidence using the ChatGPT release as an AI adoption shock. Section~\ref{sec:conclusion} discusses policy implications and future work.
% --- Clarity pass complete: 2026-03-10 ---
% --- Flow pass complete: 2026-03-11 ---
% --- Technical audit complete: 2026-03-11 ---
