% Section 2: Literature Review
\label{sec:literature}

\subsection{Global Games and Coordination Failure}

The global games framework, introduced by \citet{carlsson1993} and developed by \citet{morris1998, morris2002}, resolves coordination multiplicity through heterogeneous private signals. \citet{goldstein2005} applied this framework to bank runs. The uniqueness result depends on the information structure: \citet{hellwig2002} proved that precise public information reinstates multiplicity, and \citet{angeletos2006} showed that endogenous policy signals can similarly restore it. \citet{angeletos2007} characterised the welfare implications, establishing that more precise public information can be socially harmful because agents over-weight the common component.

When financial agents rely on the same AI model, the cross-sectional correlation of their signals rises, converting private information into a near-public common signal and activating the \citet{hellwig2002} mechanism. \citet{yang2024} demonstrates this computationally via Q-learning agents; \citet{danielsson2025} model it formally through an execution-speed channel. Neither introduces a signal correlation parameter or integrates the \citet{hellwig2002} multiplicity boundary into the formal model. The distinction between these approaches and ours is substantive. \citet{danielsson2025} derive a crisis threshold $\theta^*(\mu)$ that is linear in the fraction of AI agents $\mu$, an extensive-margin primitive. Our primitive $\rho$ captures the intensive margin of AI output homogeneity. The two primitives generate opposing risk assessments: many agents using diverse AI models (high $\mu$, low $\rho$) register as dangerous in the Danielsson-Uthemann framework but safe in ours, while few agents sharing a single dominant model (low $\mu$, high $\rho$) register as safe in theirs but fragile in ours. \citet{yang2024} models AI coordination through Q-learning agents whose crisis threshold is learned through Q-value convergence rather than derived analytically; their model has no signal correlation parameter and no connection to information acquisition or market-making channels. \citet{margaretic2014} extend the Goldstein-Pauzner framework with correlated bank fundamentals and interbank linkages, but their correlation structure operates on investment quality rather than on the signals received by agents. \citet{szkup2015} endogenise information acquisition in global games; their sufficient conditions for uniqueness may be violated when signals become correlated through a common AI source. \citet{hansen2025} find that independently queried AI agents reduce herding, consistent with the low-$\rho$ regime; the fragility mechanism activates at high $\rho$, a regime their experiment does not test.

\subsection{Information Acquisition and Price Efficiency}

\citet{grossman1980} established that informationally efficient markets are impossible when information is costly. \citet{holden1992} showed that competition among identically informed traders drives rents to zero. \citet{goldstein2015} showed that information diversity creates strategic complementarities in acquisition. \citet{bond2012} distinguished forecasting price efficiency (FPE) from revelatory price efficiency (RPE); the FPE/RPE distinction is central to evaluating AI's welfare consequences.

On the technology side, \citet{dugast2018} model how data abundance reduces informativeness through a speed-precision tradeoff in the temporal dimension; \citet{dugast2025} extend this to equilibrium data mining. \citet{farboodi2020} model long-run growth of data processing technology, and \citet{farboodi2022} document that data technology concentrates information on large firms. \citet{banerjee2018} show that greater transparency can paradoxically reduce informativeness. \citet{vives2014} established that a privately revealing equilibrium obtains when correlation in traders' valuations is not too large; AI-driven signal correlation progressively violates this condition.

The defining contrast with \citet{dugast2018} is the dimension of the tradeoff. Dugast and Foucault model the temporal choice between fast and slow signals for a single agent. The present paper models the cross-sectional dimension: $N$ agents holding the same AI signal collapse the market's effective information content regardless of each signal's precision. No existing paper models this cross-sectional dimension where correlated AI outputs collapse effective information content even though each individual signal is precise. The RPE degradation result and the FPE-RPE divergence have no predecessor in this literature. Proposition~\ref{prop:rpe} combines the \citet{holden1992} competition effect with the \citet{goldstein2015} complementarity breakdown to produce a monotonically decreasing relationship between $\rho$ and revelatory price efficiency.

\subsection{Market Making and Correlated Liquidity}

The foundational models of \citet{glosten1985} and \citet{kyle1985} established spread-setting under adverse selection. \citet{avellaneda2008} provided the stochastic control framework for optimal market making. \citet{greenwood2011} defined stock fragility through demand-side correlation from mutual fund flows. \citet{danielsson2012} formalised how common VaR constraints produce procyclical selling. \citet{brunnermeier2009} modelled the feedback between market and funding liquidity. \citet{colliard2025} model a single Q-learning market maker; \citet{cespa2025} show that opacity creates strategic complementarity in liquidity demand. Neither addresses cross-sectional correlation among multiple AI-calibrated market makers. The contrast with \citet{cespa2025} is instructive: their fragility operates through opacity and strategic complementarity in liquidity demand, while ours operates through algorithmic similarity in liquidity supply. The two mechanisms could compound in practice but are formally distinct.

We derive $N_{\text{eff}}(\rho) = N/(1+(N-1)\rho)$ as a supply-side fragility measure grounded in algorithmic similarity. The model nests \citet{danielsson2012} as the $\rho = 1$ limiting case and extends \citet{greenwood2011} from demand-side to supply-side correlation. A threshold $\rho^{**}$ exists above which no interior market-making equilibrium with finite spreads exists, analogous to the \citet{pagano1989} low-liquidity trap.

\subsection{AI and Financial Markets}

\citet{gu2020} document that diverse machine learning methods converge on the same dominant predictors, and \citet{kim2024} show that GPT-4 generates near-identical financial analyses, both documenting the empirical manifestation of rising $\rho$. \citet{calvano2020} showed that Q-learning algorithms autonomously sustain supra-competitive prices. In financial markets, \citet{dou2025a} show that RL-powered trading algorithms sustain collusive profits; \citet{dou2025b} extend this to agentic AI with planning capability. \citet{barucca2025} discuss how low-cost AI, acting as universal approximators, reshapes market efficiency in the Grossman-Stiglitz tradition, a conceptual framing that lacks the formal signal-correlation mechanism we develop. \citet{danielsson2022} provide the most comprehensive qualitative analysis of AI and systemic risk, identifying four channels but proposing no formal model.

The distinction between the reinforcement learning approach and the information-theoretic approach is fundamental. The \citet{dou2025a, dou2025b} programme models AI fragility through emergent coordination via repeated interaction. Our mechanism arises from signal homogeneity driven by common data and model architecture, captured by $\rho$. The two approaches are complementary.

\subsection{Cross-Channel Interaction and Systemic Risk}

The \citet{danielsson2022, danielsson2025} research programme is the most developed multi-channel framework: the 2022 paper identifies four qualitative channels, and the 2025 paper formalises the coordination channel. The other channels remain qualitative, and no compound fragility threshold is derived. \citet{goldstein2025} survey the theoretical forces generating market fragility but do not address AI-specific mechanisms.

No existing paper characterises the joint dynamics of coordination failure, information acquisition collapse, and correlated liquidity withdrawal through a common parameterisation $\rho$, or derives the compound threshold at which these channels amplify one another. The present paper also connects to the emerging literature on algorithmic monoculture \citep{kleinberg2021, peng2024} and AI adoption externalities \citep{gans2023}. Section~\ref{sec:model} introduces the signal structure that unifies all three channels.
% --- Clarity pass complete: 2026-03-11 ---
% --- Flow pass complete: 2026-03-11 ---
% --- Technical audit complete: 2026-03-11 ---
